---
bibliography: references.bib
---

# Methods comparison

```{r include=FALSE}

knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

# Load data
source("./scripts/0_meta_env_prep.R")

# Load DMRcate results
quant_ratioset_funnorm_filter <- readRDS(file = file.path("output", "quant_ratioset_funnorm_filter.RDS")) %>%
  .[, .@colData$cell_line == "hMSC"] %>%
  .[, .@colData$treatment != "neurosphere"]

quant_m_vals <- readRDS(file.path("output", "data_meth", "dmrcate", "quant_m_vals_hmsc.RDS"))

t_stat_timepoint <- readRDS(file.path("output", "data_meth", "dmrcate", "t_stat_timepoint.RDS"))

t_stat_heparin <- readRDS(file.path("output", "data_meth", "dmrcate", "t_stat_heparin.RDS"))

dmr_timepoint <- readRDS(file.path("output", "data_meth", "dmrcate", "dmr_timepoint.RDS"))

ranges_dmr_timepoint <- readRDS(file.path("output", "data_meth", "dmrcate", "ranges_dmr_timepoint.RDS"))

# Load limma + mCSEA results

fit_contrasts <- readRDS(
        file.path("output", "data_meth", "limma", "fit_contrasts.RDS"))

mcsea_timepoint_t_stat <- readRDS(
        file = file.path("output", "data_dmr", "mcsea", "dmr_hMSC_timepoint.RDS"))

mcsea_heparin_t_stat <- readRDS(
        file = file.path("output", "data_dmr", "mcsea", "dmr_hMSC_heparin.RDS"))

# Load direct ordering results

mcsea_timepoint_direct <- readRDS(
        file = file.path("output", "data_dmr", "mcsea_deltabeta", "dmr_hMSC_timepoint.RDS"))

mcsea_heparin_direct <- readRDS(
        file = file.path("output", "data_dmr", "mcsea_deltabeta", "dmr_hMSC_heparin.RDS"))

# Correlation between methods

df_cor_nes <- readRDS(file = file.path("output", "data_dmr", "df_cor_nes.RDS"))

```

## Intro

3 methods were tested to compare DNA methylation changes between P+5 and P+13:

1.  Calculation of CpG methylation test statistics with `limma` [@ritchie2015] by considering untreated and treated cells as one group, followed by DMR detection with DMRcate.
    -   Grouping untreated and treated cells gives us n = 2 and allows calculation of CpG-level test statistics and p-values.
    -   This method assumes that the effect of Heparin treatment does not change between passages.
        -   e.g. If Heparin increases methylation by 10% at a probe at P+5, then it still increases methylation by 10% at P+13.
        -   This assumption does not hold with hNSCs but might be more plausible with hMSCs (based on MDS clustering results).
2.  Calculation of CpG methylation test statistics with `limma` [@ritchie2015] by considering untreated and treated cells as one group, followed by permutation-based DMR detection with `mCSEA` [@martorell-marugán2019] .
    -   Grouping untreated and treated cells gives us n = 2 and allows calculation of CpG-level test statistics and p-values.
    -   This method assumes that the effect of Heparin treatment does not change between passages.
        -   e.g. If Heparin increases methylation by 10% at a probe at P+5, then it still increases methylation by 10% at P+13.
        -   This assumption does not hold with hNSCs but might be more plausible with hMSCs (based on MDS clustering results).
    -   This method applies a potentially flawed permutation-based test for DMR detection with `mCSEA`.
        -   Under the hood, `mCSEA` applies the `fgsea` algorithm [@korotkevich].
        -   `fgsea` may be flawed - it assumes that all elements in the tested list are independent of each other.
            -   In the context of gene set enrichment analysis (gsea), it assumes that expression of all genes being tested are independent of each other.
            -   In the context of DMR identification (here), it assumes that methylation of all probes being tested are independent of each other.
            -   These assumptions may not be biologically plausible.
        -   However, the original mCSEA publication demonstrated high PPV (near 100% at $\Delta\beta\geq 0.2$ , 91.29% at $\Delta \beta = 0.05$, Supplementary Table 2) in spite of theoretical flaws.
3.  Ranking of DNA methylation probe by decreasing \$\Delta \beta\$, or by decreasing levels of hypermethylation between passages, followed by permutation-based DMR detection with `mCSEA` [@martorell-marugán2019]
    -   This method follows "Option B" outlined in the `mCSEA` publication.
    -   This method does not derive CpG methylation test statistics (impossible with n = 1).
        -   For `mCSEA/fgsea`, this doesn't matter - the permutation-based tests only use the rank of CpG sites in the list, not the actual test statistic.
    -   This method applies a potentially flawed permutation-based test for DMR detection with `mCSEA`, discussed in detail above.

## Method 1: DMRcate

### CpG level results:

Between passages: 143056 differentially methylated CpG sites (FDR \< 0.05)

```{r}

t_stat_timepoint

t_stat_timepoint@ranges
```

Between treatment (heparin vs untreated): 0 differentially methylated CpG sites (FDR \< 0.05)

```{r}

t_stat_heparin

t_stat_heparin@ranges
```

### DMR results

25436 DMRs were detected, with most DMRs associated with hypomethylation.

```{r}

ranges_dmr_timepoint

ranges_dmr_timepoint@elementMetadata$meandiff %>% range()

ranges_dmr_timepoint@elementMetadata$maxdiff %>% range()
```

```{r}

datatable(as_tibble(ranges_dmr_timepoint),
          options = list(
            scrollX = TRUE
          )) %>%
  formatRound(7:12, 4)
```

## Method 2: limma + mCSEA

### CpG level results

```{r}

summary(decideTests(fit_contrasts))
```

Between passages: 143056 differentially methylated CpG sites (FDR \< 0.05)

Between treatment (heparin vs untreated): 0 differentially methylated CpG sites (FDR \< 0.05)

(Same results as DMRcate due to same method being used)

### DMR results (gene promoters only)

#### Between passages

```{r}

table(mcsea_timepoint_t_stat[["promoters"]]$padj < 0.05)
```

534 DMRs

##### All regions results

```{r}

datatable(mcsea_timepoint_t_stat[["promoters"]] %>%
            drop_na() %>%
            select(!1),
          options = list(scrollX = TRUE)) %>%
  formatRound(1:4, 4)

```

##### DMRs only

```{r}

datatable(mcsea_timepoint_t_stat[["promoters"]] %>%
            filter(padj < 0.05) %>%
            drop_na() %>%
            select(!1),
          options = list(scrollX = TRUE)) %>%
  formatRound(1:4, 4)

```

#### Between heparin and untreated

```{r}

table(mcsea_heparin_t_stat[["promoters"]]$padj < 0.05)
```

7 DMRs

##### All regions results

```{r}

datatable(mcsea_heparin_t_stat[["promoters"]] %>%
            drop_na() %>%
            select(!1),
          options = list(scrollX = TRUE)) %>%
  formatRound(1:4, 4)

```

##### DMRs only

```{r}

datatable(mcsea_heparin_t_stat[["promoters"]] %>%
            filter(padj < 0.05) %>%
            drop_na() %>%
            select(!1),
          options = list(scrollX = TRUE)) %>%
  formatRound(1:4, 4)

```

## Method 3: Direct ordering of ∆β + mCSEA

### DMR results (gene promoters only)

#### Between passages

```{r}

table(mcsea_timepoint_direct[["promoters"]]$padj < 0.05)
```

304 DMRs

##### All regions results

```{r}

datatable(mcsea_timepoint_direct[["promoters"]] %>%
            drop_na() %>%
            select(!1),
          options = list(scrollX = TRUE)) %>%
  formatRound(1:4, 4)

```

##### DMRs only

```{r}

datatable(mcsea_timepoint_direct[["promoters"]] %>%
            filter(padj < 0.05) %>%
            drop_na() %>%
            select(!1),
          options = list(scrollX = TRUE)) %>%
  formatRound(1:4, 4)

```

#### Between heparin and untreated

```{r}

table(mcsea_heparin_direct[["promoters"]]$padj < 0.05)
```

0 DMRs

##### All regions results

```{r}

datatable(mcsea_heparin_direct[["promoters"]] %>%
            drop_na() %>%
            select(!1),
          options = list(scrollX = TRUE)) %>%
  formatRound(1:4, 4)

```

## Correlation between methods 2 and 3

Method 1 (`DMRcate`) can't be directly compared with the remaining methods (`mCSEA`). `DMRcate` locates DMRs first, then annotates the identified regions against a reference genome. `mCSEA` tests on pre-defined genomic regions.

Methods 2 (`tstat`) and 3 (`beta`) derive a normalised enrichment score (NES) that represents whether probes in a region are mostly hypermethylated, or hypomethylated. There is weak correlation (p \< 0.05, r = 0.66) between the methods.

### Correlation of NES scores

```{r}

ggplot(df_cor_nes, aes(x = NEStstat, y = NESdeltabeta)) +
  geom_point() +
  geom_smooth(method = "lm") +
  annotate(
    "text",
    x = 2,
    y = 4,
    label = paste0("r = ", round(
      cor(df_cor_nes$NEStstat, df_cor_nes$NESdeltabeta), 2
    )),
    hjust = 0
  ) +
  annotate(
    "text",
    x = 2,
    y = 3.75,
    label = paste0("p = ", round(
      cor.test(df_cor_nes$NEStstat, df_cor_nes$NESdeltabeta)$p.value,
      3
    )),
    hjust = 0
  ) +
  theme_classic()

```

### Venn diagram

Out of all DMRs detected by both methods, only 119 DMRs are shared.

```{r}

ggVennDiagram(list(
  tstat = df_cor_nes[df_cor_nes$padjtstat < 0.05, ]$gene,
  beta =  df_cor_nes[df_cor_nes$padjdeltabeta < 0.05, ]$gene
))

```
