---
title: "Genomic Profiling of Drivers of hMSC in vitro Proliferation and Differentiation Potential"
format: gfm
---

# Introduction
This repository contains the source code for DNA methylation analysis in the thesis project: Genomic Profiling of Drivers of hMSC *in vitro* Proliferation and Differentiation Potential.

See the thesis here: [https://eprints.qut.edu.au/262996/](https://eprints.qut.edu.au/262996/)

# Prerequisites
Make sure the following are installed on your system:  

- Quarto
- R 4.5.1

# Reproduciblity
To reproduce the thesis figures:  

- Install all prerequisites software.
- Clone the repository; Positron is recommended for this purpose.
- Request a copy of our dataset via [minhnghia.nguyen@connect.qut.edu.au](mailto:minhnghia.nguyen@connect.qut.edu.au) and place it in the `input` folder as instructed.
- Run the following in an R terminal to download all packages:
```{r}
#| eval: False
renv::restore()
```

- Run the following in an R terminal to perform all analyses:
```{r}
#| eval: False
source(here::here("R/99_pipeline_firstrun.R"))
```
